Mastery Learning in Practice:\\ A Descriptive Analysis of Data from the Cognitive Tutor Algebra I Effectiveness Trial

### 2017-10-17 paper.Rnw
movin right along

&   \multicolumn{6}{c}{Year 1}&\multicolumn{6}{c}{Year 2}\\
<<statusTab,dependson='overallStatus',results='asis'>>=

tab1 <- dcast(subset(statusStateYr,Yr=='Yr 1'&state!='NJ'),variable~state)
tab2 <- dcast(subset(statusStateYr,Yr=='Yr 2'&state!='NJ'),variable~state)
tab <- cbind(tab1,tab2[-1])

tab$variable <- as.character(tab$variable)
cat('&')
cat(names(tab)[-1],sep='&')
cat('\\\\')
for(i in 1:nrow(tab)){
    cat(tab[i,1],round(unlist(tab[i,-1])*100),sep='&')
    cat('\\\\')
}

@@

@@

Figure \ref{fig:overallStatus} and Table \ref{tab:overallStatus} show
the proportions of worked sections in each state and study year that
ended with mastery, promotion, or reassignment, or which were the
student's final section.
In the first year, about
\Sexpr{round(mean(tab1['Mastered',c('TX','KY','MI','LA')])*100)}\% of
worked sections are mastered in every
state other than Connecticut.
Other than in Texas, about
\Sexpr{paste(round(range(tab1['Promoted',c('CT','KY','MI','LA')])*100),collapse='--')}\%
of sections end in promotion.
About \Sexpr{round(tab1['Reassigned','TX']*100)}
and \Sexpr{round(tab1['Reassigned','CT']*100)}\% of sections in Texas
and Connecticut end in reassignment, which is even rarer in the other states.

With the exception of Texas, sections in the second year tend to be
completed as they were in year 1.
In Texas, however, the percentage of sections ending in reassignment increased by a factor
of about about
\Sexpr{round(tab2['Reassigned','TX']/tab1['Reassigned','TX'])},
to about \Sexpr{round(tab2['Reassigned','TX']*100)}\%.
The proportion of Texas sections labeled ``Final'' increased as well,
most likely due to the decrease in usage.

Across states, just under
\Sexpr{round(tab1['Reassigned','Overall']*100)}\% of sections in year
1 eneded in reassignment, as did roughly
\Sexpr{round(tab2['Reassigned','Overall']*100)}\% in year 2.

\subsection{Section Mastery and Curriculum}
\begin{figure}
<<statusCur,dependson='overallStatus',fig.height=3,fig.width=6>>=
statusCurr <- statusOverall%>%filter(!is.na(Curriculum))%>% group_by(Curriculum,overall)%>%
    summarize(pgrad=mean(status=='graduated'),pfoi=mean(status=='final_or_incomplete'),
              pcp=mean(status=='changed placement'),pprom=mean(status=='promoted'))%>%
    melt(measure.vars=c('pgrad','pfoi','pcp','pprom'))
levels(statusCurr$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Graduated='pgrad')
statusCurr$Curriculum <- factor(statusCurr$Curriculum,levels=rev(levels(statusCurr$Curriculum)))
print(ggplot(statusCurr,aes(overall,value,fill=variable))+geom_col()+facet_grid(~Curriculum)+labs(y='% of Sections Worked',x='',fill='Exit Status')+scale_y_continuous(labels=percent))

@
\caption{The distributions of outcomes of worked sections, by curriculum, in the two study years.}
\label{fig:statusCur}
\end{figure}

\section{Digging Deeper into Section Reassignment}
<<cpDat,dependson=c('data','setup'),include=FALSE>>=
secLev <- data%>%filter(is.finite(status) & is.finite(timestamp) &is.finite(date))%>%
    group_by(field_id,unit,section,Year,Yr,classid2,schoolid2)%>%
    summarize(startDate=min(date),endDate=max(date),startTime=min(timestamp),endTime=max(timestamp),state=state[1],
              status=max(status),Curriculum=Curriculum[1],overall=overall[1],version=version[1])%>%
    arrange(endDate)

@@

@
\subsection{When are Students Reassigned?}

@@

\subsection{Where To?}
\begin{figure}
<<transition,fig.height=7,fig.width=6.5,dependson=c('data','setup')>>=
#### what's the next section each student works?
secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp))%>%group_by(field_id,section,unit,Curriculum,overall,year,Year,Yr,state,classid2,schoolid2)%>%summarize(time=max(timestamp),status=max(status))%>%arrange(time)%>%group_by(field_id,year,Year,Yr,state,classid2,schoolid2)%>%mutate(prevSec=c(NA,section[-n()]),prevStatus=c(NA,status[-n()]),prevUnit=c(NA,unit[-n()]),nextUnit=c(unit[-1],NA))

@@

trans <- with(secOrderCP,table(unitName,nextUnitName))

@@

@@

\begin{figure}
<<vcs,fig.height=7,fig.width=6.5,dependson=c('data','setup')>>=
load('vcMods.RData')
library(lme4)

@@

\begin{figure}
<<xirtCP,fig.height=7,fig.width=6.5,dependson=c('data','setup')>>=
load('cpXirtMods.RData')

xirtDat <- data%>%filter(!is.na(status) & !is.na(xirt))%>%
    group_by(field_id,classid2,schoolid2,xirt,year,spec_gifted,spec_speced,state,unit,section)%>%
    summarize(status=max(status,na.rm=TRUE),nprob=n())%>%
    group_by(field_id,classid2,schoolid2,xirt,year,spec_gifted,spec_speced,state)%>%
    summarize(nprob=sum(nprob),nsec=n_distinct(section,unit),propMast=mean(status=='graduated'),
              propCP=mean(status=='changed placement'))%>%arrange(state)

xirtDat$fitted <- NA
for(st in names(cpModSt)) xirtDat$fitted[xirtDat$state==st] <-
                                  predict(cpModSt[[st]],xirtDat[xirtDat$state==st,],
                                          allow.new.levels=TRUE,re.form=~0,type='response')

xirtDat$fitted[xirtDat$state=='Overall'] <-
    predict(cpModSimp,xirtDat[xirtDat$state=='Overall',],
            allow.new.levels=TRUE,re.form=~0,type='response')

ggplot(subset(xirtDat,state!='NJ'),aes(xirt,propCP,size=nsec))+geom_point()+geom_line(aes(xirt,fitted),col='red',size=1)+facet_grid(year~state)+coord_cartesian(ylim=c(0,0.05))

>>>>

&   \multicolumn{6}{c}{Year 1}&\multicolumn{6}{c}{Year 2}\\
<<statusTab,dependson='overallStatus',results='asis'>>=
statusStateYr2 <- addOverallState(statusOverall)
statusStateYr2a <- statusStateYr2%>%group_by(state,Yr)%>%
    summarize(pgrad=sum(status=='graduated'),pfoi=sum(status=='final_or_incomplete'),
              pcp=sum(status=='changed placement'),pprom=sum(status=='promoted'))%>%
    melt(measure.vars=c('pgrad','pfoi','pcp','pprom'))
statusStateYr2b <- statusStateYr2%>%
    summarize(pgrad=mean(status=='graduated'),pfoi=mean(status=='final_or_incomplete'),
              pcp=mean(status=='changed placement'),pprom=mean(status=='promoted'))%>%
    melt(measure.vars=c('pgrad','pfoi','pcp','pprom'))

levels(statusStateYr2a$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Mastered='pgrad')
levels(statusStateYr2b$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Mastered='pgrad')

tab1 <- dcast(subset(statusStateYr2a,Yr=='Yr 1'&state!='NJ'),variable~state)
tab2 <- dcast(subset(statusStateYr2a,Yr=='Yr 2'&state!='NJ'),variable~state)
tab <- cbind(tab1,tab2[-1])

tab1b <- dcast(subset(statusStateYr2b,Yr=='Yr 1'&state!='NJ'),variable~state)
tab2b <- dcast(subset(statusStateYr2b,Yr=='Yr 2'&state!='NJ'),variable~state)

tab$variable <- as.character(tab$variable)
cat('&')
cat(names(tab)[-1],sep='&')
cat('\\\\')
for(i in 1:nrow(tab)){
    cat(tab[i,1],round(unlist(tab[i,-1])),sep='&')
    cat('\\\\')
}

@@

rownames(tab1b) <- rownames(tab2b) <- tab1b$variable
tab1b <- as.matrix(tab1b[,-1])
tab2b <- as.matrix(tab2b[,-1])

@@

Figure \ref{fig:overallStatus} and Table \ref{tab:overallStatus} show
the proportions of worked sections in each state and study year that
ended with mastery, promotion, or reassignment, or which were the
student's final section.
In the first year, about
\Sexpr{round(mean(tab1b['Mastered',c('TX','KY','MI','LA')])*100)}\% of
worked sections are mastered in every
state other than Connecticut.
Other than in Texas, about
\Sexpr{paste(round(range(tab1b['Promoted',c('CT','KY','MI','LA')])*100),collapse='--')}\%
of sections end in promotion.
About \Sexpr{round(tab1b['Reassigned','TX']*100)}
and \Sexpr{round(tab1b['Reassigned','CT']*100)}\% of sections in Texas
and Connecticut end in reassignment, which is even rarer in the other states.

With the exception of Texas, sections in the second year tend to be
completed as they were in year 1.
In Texas, however, the percentage of sections ending in reassignment increased by a factor
of about about
\Sexpr{round(tab2b['Reassigned','TX']/tab1b['Reassigned','TX'])},
to about \Sexpr{round(tab2b['Reassigned','TX']*100)}\%.
The proportion of Texas sections labeled ``Final'' increased as well,
most likely due to the decrease in usage.

Across states, just under
\Sexpr{round(tab1b['Reassigned','Overall']*100)}\% of sections in year
1 eneded in reassignment, as did roughly
\Sexpr{round(tab2b['Reassigned','Overall']*100)}\% in year 2.

\subsection{Section Mastery and Curriculum}
\begin{figure}
<<statusCur,dependson='overallStatus',fig.height=3,fig.width=6>>=
statusOverall$Curr2 <- with(statusOverall,
                        ifelse(Curriculum=='Algebra I',
                                     ifelse(overall=='Standard','Algebra I','Algebra I (Cust.)'),
                                     as.character(Curriculum)))
statusCurr <- statusOverall%>%filter(!is.na(Curriculum))%>% group_by(Curr2,Yr)%>%
    summarize(pgrad=mean(status=='graduated'),pfoi=mean(status=='final_or_incomplete'),
              pcp=mean(status=='changed placement'),pprom=mean(status=='promoted'))%>%
    melt(measure.vars=c('pgrad','pfoi','pcp','pprom'))
levels(statusCurr$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Graduated='pgrad')
statusCurr$Curr2 <- factor(statusCurr$Curr2,levels=c('Bridge-to-Algebra','Algebra I','Algebra I (Cust.)','>Algebra I'))
print(ggplot(statusCurr,aes(Yr,value,fill=variable))+geom_col()+facet_grid(~Curr2)+
      labs(y='% of Sections Worked',x='',fill='Exit Status')+scale_y_continuous(labels=percent))

@
\caption{The distributions of outcomes of worked sections, by
  curriculum, in the two study years. (There were no Bridge-to-Algebra
sections in customized curricula in our dataset.)}
\label{fig:statusCur}
\end{figure}

A well-designed curriculum, can, in theory, play an important role in
students' attainment of mastery.
Students who work on appropriate problems that build on their current
set of skills should be more likely to master new skills than students
working on problems above their level.
What role did curriculum play in mastery during the CTAI effectiveness trial?

Figure \ref{fig:statusCur} shows the proportions of worked sections
that were mastered or ended in promotion, reassignment, or finality, in
standard and customized versions of each CT curriculum.
Mastery proportions do, indeed, depend on curriculum.
Specifically, students mastered sections from more advanced curriculua
less frequently.
Sections from the most basic curriculum, Bridge to Algebra, were
mastered
\Sexpr{round(statusCurr[['value']][statusCurr[['Curriculum']]=='Bridge-to-Algebra'
  & statusCurr[['variable']]=='Graduated']*100)}\% of the time;
those from Algebra I were mastered
\Sexpr{round(
  mean(statusOverall$status[statusOverall$Curriculum=='Algebra I']=='graduated',na.rm=T)*100)}\%
of the time, and those from more advanced curriculua were mastered
at a rate of
\Sexpr{round(
  mean(statusOverall$status[statusOverall$Curriculum=='>Algebra I']=='graduated',na.rm=T)*100)}\%.
This is unsurprising, since more advanced curriculua may be expected
to be more challenging.
However, it may suggest that some students studying advanced topics
would fair better in more standard curricula.

Algebra I sections from customized curricula tended to end in
reassignment more often than sections from the standard Algebra I
curriculum (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' &
  variable=='Reassigned')[['value']])}\% vs.
\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Reassigned')[['value']])}\%).
This may an overall skepticism towards the Carnegie Learning standards
among certain schools and teachers.

% A somewhat lower proportion of Algebra I sections from customized
% curricula were mastered
% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' & variable=='Graduated')[['value']])}\%)
% than from the standard Algebra I curriculum
% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Graduated')[['value']])}\%).
% This difference must be interpreted with caution, since a greater
% proportion of sections of the customized curriculum ended in
% reassignment

% or were final
% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' &   variable=='Final')[['value']])}\% vs.
% \Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Final')[['value']])}\%)
% , as well.
% The proportion of sections ending in promotion was almost identical:
% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' &variable=='Promoted')[['value']])}\% vs.
% \Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Promoted')[['value']])}\%).

\section{Digging Deeper into Section Reassignment}
The proportion of worked sections in our dataset ending in
reassignment was small.
Nevertheless, since reassignement represents the only mechanism by
which individual teachers can affect their students' progress through
the Cognitive Tutor, exploring patterns of reassignment can provide
insight into how CT was used.

<<cpDat,dependson=c('data','setup'),include=FALSE>>=
secLev <- data%>%filter(is.finite(status) & is.finite(timestamp) &is.finite(date))%>%
    group_by(field_id,unit,section,Year,Yr,classid2,schoolid2)%>%
    summarize(startDate=min(date),endDate=max(date),startTime=min(timestamp),endTime=max(timestamp),state=state[1],
              status=max(status),Curriculum=Curriculum[1],overall=overall[1],version=version[1])%>%
    arrange(endDate)

@@

\subsection{How Do Reassignment Patterns Vary?}
Teachers alone control reassignment.
Nevertheless, the factors influencing student reassignment vary at a
number of levels.
For instance, state and district standards may prod teachers into
reassigning students to particular units.
Some prinicpals may encourage teachers to adhere to the official
curriculum and avoid reassignment.
Some students may be more prone to reassignment than others.
Certain units in the CTAI curriculum may be harder than others,
causing students to tarry and teachers to reassign.
Finally, a host of other factors, at these levels and others, may spur reassignment.

Figure \ref{fig:vc}

\begin{figure}
<<vcs,fig.height=3,fig.width=6.5,dependson=c('data','setup')>>=
load('vcMods.RData')
library(lme4)

@
\caption{Results from a set of eight multilevel logistic regressions
  predicting section reassignment. For each year, in the entire sample (``Overall'')
  and in the three states with the largest numbers of reassignments (Texas,
  Kentucky, and Michigan), we regressed a binary variable indicating
  whether a section ended in reassignment on random intercepts for
  school, class, student, and unit, and in the overall case, for state
  as well, and recorded their variance. The residual variance was set
  as the variance of the standard logistic distribution,
  $\pi/3$. These bar charts give the proportion of the total variance
  attributable to each random effect.}
\label{fig:vc}
\end{figure}

\subsection{When are Students Reassigned?}

@@

\subsection{Where To?}

<<secOrder,include=FALSE,dependson=c('data','setup')>>=
#### what's the next section each student works?
secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp))%>%group_by(field_id,section,unit,Curriculum,overall,year,Year,Yr,state,classid2,schoolid2)%>%summarize(time=max(timestamp),status=max(status))%>%arrange(time)%>%group_by(field_id,year,Year,Yr,state,classid2,schoolid2)%>%mutate(prevSec=c(NA,section[-n()]),prevStatus=c(NA,status[-n()]),prevUnit=c(NA,unit[-n()]),nextUnit=c(unit[-1],NA))

@@

\begin{figure}
<<transition1,fig.height=3,fig.width=6.5,dependson='secOrder'>>=

trans <- with(filter(secOrderCP,year==1),table(unitName,nextUnitName))

@@

propTotCP <- sum(trans)/sum(secOrderCP$year==1)
#transitionPlot(trans)

\begin{figure}
<<transition2,fig.height=7,fig.width=6.5,dependson='secOrder'>>=

trans <- with(filter(secOrderCP,year==2),table(unitName,nextUnitName))

propTotCP <- sum(trans)/sum(secOrderCP$year==2)

@@

\subsection{Does Reassignment Depend on Student Ability?}

<<xirtMods,include=FALSE,dependson=c('data','setup')>>=
## don't worry about nesting, etc--we aren't reporting standard errors anyway
cpDat <- data%>%filter(!is.na(status))%>%group_by(field_id,unit,section,state,year,classid2,schoolid2)%>%
    summarize(cp=max(status)=='changed placement')

### get imputed xirt
##
h1 <- read.csv('~/Box Sync/CT/data/RANDstudyData/H1_algebra_rcal_20121119_fieldid.csv')
h2 <- read.csv('~/Box Sync/CT/data/RANDstudyData/H2_algebra_rcal_20121119_fieldid.csv')

h2 <- h2[!h2$field_id%in%h1$field_id,]

stud <- rbind(h1[,c(174:194)],h2[,c(176:196)])
stud <- stud[stud$field_id%in%cpDat$field_id,]
stud <- stud[order(stud$field_id),]
stud$xirt <- rowMeans(stud[,grep('Exirt2_',names(stud))],na.rm=TRUE)

cpDat$xirt <- stud$xirt[match(cpDat$field_id,stud$field_id)]

cpModSimp <- glm(cp~splines::ns(xirt,5)*year,data=cpDat)

cpModSt <- list()
for(st in c('TX','KY','MI','LA','CT')){
    cpModSt[[st]] <- update(cpModSimp,data=subset(cpDat,state==st))
}

\begin{figure}
<<xirtCP,fig.height=7,fig.width=6.5,dependson=c('data','setup','xirtMods')>>=
#load('cpXirtMods.RData')

xirtDat <- cpDat%>%group_by(field_id,state,year,classid2,schoolid2,xirt)%>%
    summarize(nsec=n_distinct(section,unit),propCP=mean(cp,na.rm=TRUE))

for(st in names(cpModSt)) xirtDat$fitted[xirtDat$state==st] <-
                                  predict(cpModSt[[st]],xirtDat[xirtDat$state==st,],
                                         type='response')

xirtDat$fitted[xirtDat$state=='Overall'] <-
    predict(cpModSimp,xirtDat[xirtDat$state=='Overall',],
            type='response')

ggplot(subset(xirtDat,state!='NJ'),aes(xirt,propCP,size=nsec))+geom_point()+geom_line(aes(xirt,fitted),col='red',size=1)+facet_grid(year~state)+coord_cartesian(ylim=c(0,0.5))

@@

\begin{figure}
<<cp-xirtClassroom1,dependson='xirtCP',fig.height=7,fig.width=6.5>>=
bbb2 <- data%>%filter(status=='changed placement')%>%group_by(classid2,year,state)%>%summarize(ncp=n_distinct(field_id,section,unit),nstud=n_distinct(field_id),nsec=n_distinct(section))

cpcls1 <- xirtDat%>%
    filter(classid2%in%bbb2$classid2[bbb2$year==1 & bbb2$ncp>=20] & state!='Overall')%>%
    arrange(state)
cpcls1$classid2 <- factor(cpcls1$classid2,levels=unique(cpcls1$classid2))

cpcls1 <- cpcls1%>%group_by(classid2)%>%mutate(maxCP=max(propCP,na.rm=TRUE),propCP=propCP/maxCP)%>%ungroup()

ggplot(cpcls1,aes(xirt,propCP,size=nsec,color=state))+
    geom_point()+geom_smooth(size=1,method='lm')+facet_wrap('classid2')+scale_size_continuous(range=c(0.2,1))+
    ylim(0,1)

\begin{figure}
<<cp-xirtClassroom2,dependson='cp-xirtClassroom1',fig.height=7,fig.width=6.5>>=

cpcls2 <- xirtDat%>%
    filter(classid2%in%bbb2$classid2[bbb2$year==2 & bbb2$ncp>=20] & state!='Overall')%>%
    arrange(state)
cpcls2$classid2 <- factor(cpcls2$classid2,levels=unique(cpcls2$classid2))

cpcls2 <- cpcls2%>%group_by(classid2)%>%mutate(maxCP=max(propCP,na.rm=TRUE),propCP=propCP/maxCP)%>%ungroup()

ggplot(cpcls2,aes(xirt,propCP,size=nsec,color=state))+
    geom_point()+geom_smooth(size=1,method='lm')+facet_wrap('classid2')+scale_size_continuous(range=c(0.2,1))+
    ylim(0,1)
